---
title: "Inspecting and Cleaning Data in R"
output: pdf_document
date: "2023-02-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)

NCbirths <- read_csv("https://wilkelab.org/DSC385/datasets/NCbirths.csv")

head(NCbirths)

```

## Cursory Inspection

The `MomAge` variable is represented as a double when it should
be an int and `Gained` is representing numbers as characters. There is probably missing data in these columns. The `read_csv()` function misinterpreted these values and converted everything to a char to compensate.

```{r}
NCbirths |>
  pull(Gained)
```

You have to be explicit with `read_csv` about how to handle missing data.

```{r}

NCbirths <- read_csv("https://wilkelab.org/DSC385/datasets/NCbirths.csv", na = c("", "NA", "#N/A"))
```

R is powerful for data analysis because it has a built-in concept for missing data (NA).

### Detailed Quality Checks: Look at all column names

```{r}
names(NCbirths)
```

What is the difference between RaceMom and MomRace. Look at the data dictionary for this dataset. RaceMom encodes numerical values to various races from 1-8, and MomRace uses strings to explicitly specify the race for a particular person. Encoding numerical values to these sorts of columns can lead nonsensical things like calculating the mean. Be very careful you know what the numbers in a column represent. Read the data dictionary if one is supplied. If a data dictionary is not supplied, maybe find a different, better dataset to answer your question.

### Detailed Quality Checks: Look at every data column

```{r}
summary(NCbirths$ID)

unique(NCbirths$ID)
```

Histograms provide a good sanity check when inspecting columns.

```{r}
summary(NCbirths$Plural)

ggplot(NCbirths, aes(Plural)) +
  geom_histogram()
```

```{r}
summary(NCbirths$Weeks)

ggplot(NCbirths, aes(Weeks)) +
  geom_histogram()
```

What is the meaning of -999 weeks?

This is another representation of missing data.

```{r}
library(naniar) # library for handling missing data

NCbirths <- read_csv(
  "https://wilkelab.org/DSC385/datasets/NCbirths.csv",
  na = c("", "NA", "#NA")
) |>
  replace_with_na_at("Weeks", ~.x < 0) # only change values in the Weeks column.
```

This column should be fixed now.

```{r}
summary(NCbirths$Weeks)

ggplot(NCbirths, aes(Weeks)) +
  geom_histogram()
```

### Consider recoding categorical variables

Sex of the baby is encoded as 1/2.

```{r}
NCbirths_clean <- NCbirths |>
  mutate(
    Sex = if_else(Sex == 1, "male", "female")
  )

NCbirths_clean |>
  pull(Sex) |>
  head()
```
### Recode multiple categories with `case_when()`

```{r}
NCbirths_clean |>
  mutate(
    RaceMom = case_when(
      RaceMom == 1 ~ "white",
      RaceMom == 2 ~ "black",
      RaceMom == 3 ~ "American Indian",
      RaceMom == 4 ~ "Chinese",
      RaceMom == 5 ~ "Japanese",
      RaceMom == 6 ~ "Hawaiian",
      RaceMom == 7 ~ "Filipino",
      RaceMom == 8 ~ "Other Asian or Pacific Islander",
      TRUE ~ NA_character_ # fallback if new data is added with missing values or new categories
    )
  )
```
### Make pairwise scatter plots

After you have inspected all of the columns in the dataset, you should make pairwise scatterplots to learn how the different variables relate to each other.

```{r}
ggplot(NCbirths_clean, aes(Plural, Sex, color=Sex)) +
  geom_point(position = "jitter", size = 0.5)
```

```{r}
ggplot(NCbirths_clean, aes(Plural, MomAge, color=Sex)) +
  geom_point(position = "jitter", size = 0.5)
```

Ideally make scatterplots for every variable against every other variable. This is the best way to understand your dataset. Make hypotheses as you go through this process. Can the dataset answer your hypotheses?